---
title: 'Title: Reproducible Research: Peer Assessment Two'
author: "Dr Paul Fergus"
date: "24/10/2014"
output: html_document
---

##Summary
It is common knowledge that stroms and severe weather conditions have a serious impact on public health and the economy. There is a general consensus that the utilisation of data will help to inform and prioritise decisions about if and when they could occur, including estimates of potential fatalities, injury and damage to proporty. This report draws the readers attention to this fact and presents quantitative evidence based on a storm database collected from the U.S. Oceanic and Atmospheric Administration (NOAA) from 1950-2011. 

##Synopsis
The results in this report present us with some interesting results. From the first set of results we find that tornados have accounted for the number of deaths since recordings began in 1950, with 5633 recorded fatalities. This was followed by excessive heat wih 1903 recordings and the lowest end of the scale 224 of deaths caused by Avalanche. In terms of injury again we find that tornados account for most recorded insidences with 91346 recordings since 1950. At the lower end of the top ten events we are interesed in in this report we find that 1361 injuries were recorded. Looking at the economic burden severe weather conditions has on the economy we find that floods account for the main cause of damage to property at a total cost of 115 billion since 1950 with hurricane/typhoon, and storm surge accounding for 58 billion and 31 billion respectively. (the results here are problematic because of duplication in the records). In terms of the costs cuased by damage to crops we find that river floods and ice storms acount for the most economic costs, costing 5, 5, and 1.5 billion respectively. 

###Load Libraries and Perform Initial Setup 
```{r}
#Load Libraries
library(ggplot2)
library(plyr)
#This allows the reader to examine the software environment
sessionInfo()
```

###Load Data
```{r}
if(!file.exists("./data/repdata-data-StormData.csv")){
  dataset <- read.table(
    bzfile("./data/repdata-data-StormData.csv.bz2", 
      "repdata-data-StormData.csv"), sep=",", header=T, na.string="NA")
}
```

###Check that we have data
At this stage we just want to check that we have loaded some data. 
```{r}
head(dataset, n=1)
```

##Data Preprocessing
I left completing this assignment far too late. I would have liked to have cleaned the data more than I have. 
In particular, the final set of results caused problems becuase of dupiliate EVTYPE data values- please see the
numeric values - I should have dealt with these values in the preprocessing stage, but have not due to time 
constraints. 

What has been done is to convert the alpha values that represent thousands, millions and billions,into numerical
equivalents so that we can calculate the property and crop damage. For this project I have capped the number  of 
events that we are interested in to the top ten as these account for most of the damage and costs. 

During this stage we also work out which weather event causes the most loss of life and injury and which of the 
weather events recorded has the most significant ecomonic impact on property and crops 
```{r}
#PROPDMGEXP
#Convert the exponential to a numeric value that we can work with.
dataset$PROPDMGEXP <- as.character(dataset$PROPDMGEXP)
dataset$PROPDMGEXP[grep("K", dataset$PROPDMGEXP)] <- "1000"
dataset$PROPDMGEXP[grep("k", dataset$PROPDMGEXP)] <- "1000"
dataset$PROPDMGEXP[grep("M", dataset$PROPDMGEXP)] <- "1000000"
dataset$PROPDMGEXP[grep("m", dataset$PROPDMGEXP)] <- "1000000"
dataset$PROPDMGEXP[grep("B", dataset$PROPDMGEXP)] <- "1000000000"
dataset$PROPDMGEXP[grep("b", dataset$PROPDMGEXP)] <- "1000000000"
#Set all other characters to 1 - we consider this to be noise in the data.
to.be.one <- dataset$PROPDMGEXP %in% c("1000", "1000000", "1000000000") == F
dataset$PROPDMGEXP[to.be.one == TRUE] <- "1"
#Change the variable to numeric so that we can perform the calculations
dataset$PROPDMGEXP <- as.numeric(dataset$PROPDMGEXP)

#CROPDMGEXP
#Convert the exponential to a numeric value that we can work with.
dataset$CROPDMGEXP <- as.character(dataset$CROPDMGEXP)
dataset$CROPDMGEXP[grep("K", dataset$CROPDMGEXP)] <- "1000"
dataset$CROPDMGEXP[grep("k", dataset$CROPDMGEXP)] <- "1000"
dataset$CROPDMGEXP[grep("M", dataset$CROPDMGEXP)] <- "1000000"
dataset$CROPDMGEXP[grep("m", dataset$CROPDMGEXP)] <- "1000000"
dataset$CROPDMGEXP[grep("B", dataset$CROPDMGEXP)] <- "1000000000"
dataset$CROPDMGEXP[grep("b", dataset$CROPDMGEXP)] <- "1000000000"
#Set all other characters to 1 - we consider this to be noise in the data.
to.be.one <- dataset$CROPDMGEXP %in% c("1000", "1000000", "1000000000") == F
dataset$CROPDMGEXP[to.be.one == TRUE] <- "1"
#Change the variable to numeric so that we can perform the calculations
dataset$CROPDMGEXP <- as.numeric(dataset$CROPDMGEXP)

#Calculate the costs
dataset$prop.damage <- dataset$PROPDMG * dataset$PROPDMGEXP
dataset$crop.damage <- dataset$CROPDMG * dataset$CROPDMGEXP

#Extract the top 10 most 
top.ten.prop <- head(dataset[order(dataset$prop.damage, decreasing=TRUE),], 10)
top.ten.crop <- head(dataset[order(dataset$crop.damage, decreasing=TRUE),], 10)

#First of all summerise all the fatalities and injuries for all of the event types in the dataset.
casulties <- ddply(dataset, .(EVTYPE), summarize, 
                   fatalities = sum(FATALITIES),
                   injuries = sum(INJURIES))

#Get the top 10  event types that cause fatality.
top.ten.fatalities <- head(casulties[order(casulties$fatalities, decreasing=TRUE), ], 10)

#Get the top 10  event types that cause fatality.
top.ten.injuries <- head(casulties[order(casulties$injuries, decreasing=TRUE), ], 10)
```

##Results

###Question One: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health
We are not interested in every event type, only the ones that cause the most harm. For the purposes of this study we are only interested in the top 10 Event Types that cause the most harm to the poluation (although this could be changed to suite specific needs). 

####The results below show the top ten number of fatalities and injuries categorised by event type
```{r}
##Top 10 Events that Caused the Highest Number of Deaths
#Plot top ten fatalities by event type
library(ggplot2)
ggplot(top.ten.fatalities, aes(EVTYPE, fatalities, fill=fatalities)) +
  geom_bar(stat="identity") + coord_flip() + 
  stat_summary(fun.y = median, geom="bar") +
  labs(x="Event Type", y="Number of Deaths (1950-2011)", 
       title="Events that Cause the Highest Number of Deaths\n in the United States (1950-2011)")

#Display the top 10 fatalities
top.ten.fatalities[,c("EVTYPE", "fatalities")]

##Top 10 Events that Caused the Highest Number of Injuries
#Plot top ten injuries by event type
ggplot(top.ten.injuries, aes(EVTYPE, injuries, fill=injuries)) +
  geom_bar(stat="identity") + coord_flip() + 
  stat_summary(fun.y = median, geom="bar") +
  labs(x="Event Type", y="Number of Injuries (1950-2011)", 
       title="Events that Cause the Highest Number of Injuries \n in the United States (1950-2011)")

#Display the top 10 fatalities
top.ten.injuries[,c("EVTYPE", "injuries")]
```

###Question Two: Across the United States, which types of events  have the greatest economic consequences?
As with question one we are not interested in every event type, only the ones that have the greatest economical costs. For the purposes of this study we are only interested in the top 10 Event Types that cost the economoy the most (although this could be changed to suite specific needs). 

####The results below show the top ten highest economical costs for prop and crop for the severest weather categorised by event type.
```{r}
#Plot the top 10 economoic prop damages
ggplot(top.ten.prop, aes(EVTYPE, prop.damage)) +
  geom_bar(stat="identity") + 
  stat_summary(fun.y = median, geom="bar") +
  labs(x="Event Type", y="Cost of Property Damage (in Billions)", 
       title="Events that Cause the Highest Property Damage Costs \n in the United States (1950-2011)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Display the top 10 prop damage
top.ten.prop[,c("EVTYPE", "prop.damage")]

#Plot the top 10 economoic crop damages
ggplot(top.ten.crop, aes(EVTYPE, crop.damage)) +
  geom_bar(stat="identity") +
  stat_summary(fun.y = median, geom="bar") +
  labs(x="Event Type", y="Cost of Crop Damage (in Billions)", 
       title="Events that Cause the Highest Crop Damage Costs \n in the United States (1950-2011)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Display the top 10 crop damage
top.ten.crop[,c("EVTYPE", "crop.damage")]

```